Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-10 paper-new/towards-relational-model.tex
revise sections on relational model

At this point we might try to weaken from our initial attempt at
providing a poset semantics to a semantics where types are equipped
with merely a reflexive relation, and continue using the
step-insensitive ordering. However some level of transitivity is
absolutely essential to providing compositional reasoning, and is used
pervasively in the New-Licata development.
%
The reason that it is essential has to do with the details of the
graduality proof, so first we expand on how we prove graduality
compositionally. First of all, we need to extend our compositional
semantics of types and terms to also a compositional semantics of
\emph{type precision derivations} and \emph{term precision}.
%
Sticking to a poset-based semantics, if our value types are posets, we
then want our precision derivations $c : A \ltdyn A'$ to be
\emph{poset relations} on $A,A'$, that is relations between the
underlying sets of $A,A'$ that are \emph{downward-closed} in $A$ and
\emph{upper-closed} in $A'$. The obvious example of such a relation is
the poset ordering itself $\leq_A : A \rel A'$...\max{TODO: Eric pick
  up from here. Go into the squares for casts and how representability
  gives us a compositional approach to proving them}

\eric{Read this}
The lack of transitivity presents a major barrier towards giving a compositional
semantics to gradual typing: if our relations are not transitive, then we cannot
compose squares horizontally. But we have argued above that horizontal
composition is essential for modelling the axioms of gradual typing in a
syntax-independent manner.

%Because it is convenient to make use of transitive reasoning in proving graduality, ...
We therefore seek to formulate an ordering relation on $\li A \times \li A'$
that \emph{is} transitive. The theorem above tells us, however, that any such
relation must either not be a congruence with respect to $\theta$, or must not
be closed under delay.
\footnote{Technically, we only need to give up closure under delay on \emph{one}
side, as the above theorem requires $R$ to be closed on both sides. We could
define two separate ``step-semi-sensitive'' relations on $\li A \times \li A'$,
each closed under delays on the left and right respectively. The denotation of
term precision would then involve the conjunction of those relations. We do not
pursue this approach in this paper, because (1) it is not clear if the
step-semi-sensitive relations would be transitive, and (2) we would need to
reason about the two ordering relations separately for every pair of terms.}
%
We cannot forego the property of being a $\theta$-congruence, as without this we
would not be able to prove properties of the relation using \lob-induction,
e.g., that $\ext{f}{}$ is a \emph{monotone} function with respect to the
ordering.
%and moreover this would mean that $\theta$ would not be a monotone function.

Thus, we choose to sacrifice closure under delays. That is, we will define an
ordering relation that requires terms to be in ``lock-step''. In order for two
computations to be related in this ordering, they must have the same stepping
behavior (or the left-hand side must be an error). To deal with terms that take
differing numbers of steps, we then define a separate relation called \emph{weak
bisimilarity} that relates terms that are extensionally equal and may differ by
a finite number of delays. Then, the semantics of the error ordering for the
guarded lift monad will involve a combination of these two relations: the
``closure'' of the lock-step error ordering under weak bisimilarity on both
sides.
% Although the combined relation will not be transitive (for the same reason that
% $\semltbad$ is not transitive), this...
This decomposition has the advantage that we can recover some transitive
reasoning and push much of the reasoning about stepping to the margins of the
development.

@@

As mentioned, to recover transitive reasoning we will decompose the error
ordering on the guarded lift monad into two separate relations. The first is the
\emph{lock-step error-ordering} (also called the ``step-sensitive error
ordering''), the idea being that two computations $l$ and $l'$ are related if
they are in lock-step with regard to their intensional behavior, up to $l$
erroring. This will be the partial ordering with which we equip the guarded lift monad.
Figure \ref{fig:lock-step-error-ordering} gives the definition.

\begin{figure}
    \fbox{$l_1 \ltls l_2$}
    \begin{align*}
        &\eta\, x \ltls \eta\, y \text{ if } 
            x \mathbin{R} y \\
        %		
        &\mho \ltls l' \\
        %
        &\theta\, \tilde{l} \ltls \theta\, \tilde{l'} \text{ if } 
            \later_t (\tilde{l}_t \ltls \tilde{l'}_t)
    \end{align*}
    \caption{The lock-step error ordering.}
    \label{fig:lock-step-error-ordering}
\end{figure}

When both sides are $\eta$, then we check that the returned values are related
in $R$. The error term $\mho$ is below everything. Lastly, if both sides step
(i.e., are a $\theta$) then we compare the resulting computations one time step
later.
%
It is straightforward to prove using \lob-induction that this relation is
reflexive, transitive and anti-symmetric given that the underyling relation $R$
has those properties. The lock-step ordering is therefore the partial ordering
we will associate with $\li X$.
%
We similarly define a heterogeneous version of this ordering between the lifts
of two different types $X$ and $Y$, parameterized by a relation $R$ between $X$
and $Y$.

The second relation we define on the guarded lift monad is \emph{weak bisimilarity}.
% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
% written $l \bisim l'$. 
Intuitively, we say $l \bisim l'$ if they are the same ``up to delays''.
%
Weak bisimilarity on $\li X$ is parameterized by an underlying relation $R$ on $X$
and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.

@@

% This issue of non-transitivity will be resolved when we consider the relation's \emph{globalization}.

@@

% Indeed, we will need to weaken our abstract notion of categorical model to no
% longer require transitivity of term precision. On the other hand, we now have
% the ability to apply transitive reasoning when working with the lock-step
% ordering. This allows us to carry out complex proofs compositionally, a key goal
% of our framework.

\subsection{Modeling Type and Term Precision}
\max{TODO: figure out where this stuff goes}
% We will model type precision $c : A \ltdyn A'$ as a relation between the sets
% $\sem{A}$ and $\sem{A'}$.

We likewise define a notion of \emph{error domain relation} between error
domains $B$ and $B'$. An error domain relation $d : B \rel B'$ is a monotone
relation on the underlying sets that preserves error and respects $\theta$. More
formally:

\begin{definition}
  Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
  \begin{enumerate}
     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.
     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
  \end{enumerate}
\end{definition}

Because of a technical restriction involving the interaction between
propositional truncation and the later modality, we cannot define the
composition of error domain relations as simply the composition of their
underlying predomain relations. Instead, we define composition of error domain
relations $d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ to be the least
relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
Specifically, it is defined inductively by the following rules:
%
\begin{mathpar}
    \inferrule*[right = Comp]
    {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
    {b_1 \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = DnClosed]
    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
    {b_1' \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = UpClosed]
    {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
    {b_1 \mathbin{d \relcomp d'} b_3'}

    \inferrule*[right = PresErr]
    { }
    {\mho_{B_1} \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = PresTheta]
    {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
    {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
\end{mathpar}

% That is,
% $(\mho_B, y) \in d$ for all $y : B'$ and if $\laterhs_t ((\tilde{x}_t,
% \tilde{y}_t) \in d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in
% d$.

The functors $U$, $F$, $\times$, and $\arr$ introduced in the previous
section extend to relations. The functor $U$ simply takes the underlying
monotone relation of an error domain relation. Given a relation $c_1 : A_1 \rel
A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2)
\rel (A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$
if and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
and respects $\theta$ follows from the corresponding properties for $d$.
%
We we will postpone the description of the action of $F$ on relations to later
in this section.

\subsubsection{Modelling Term Precision}\label{sec:modeling-term-precision}
To model term precision, we need the notion of \emph{square}, which specifies a
relation between two monotone functions. The definition is as follows. Let $A_i,
A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o :
A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g : A_i'
\to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x
: A_i$ and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
visualize this situation as the following square:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_i} \& {A_i'} \\
	{A_o} \& {A_o'}
	\arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["f"', from=1-1, to=2-1]
	\arrow["g", from=1-2, to=2-2]
	\arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]

Having defined squares, we now discuss the interpretation of the cast rules
UpL/UpR/DnL/DnR. These rules specify a relationship between the semantics of
type precision and the corresponding casts (see Section \ref{sec:GTLC}). In
particular, they specify that the relations are \emph{representable} by the
casts. The UpL and UpR rules refer to upcasts, and since upcasts are pure, these
rules apply to \emph{value} relations $c$. The DnL and DnR rules refer to
downcasts, and since downcasts are effectful, these rules apply to
\emph{computation} relations $d$.

As an example, consider the UpL rule. It states that if $c : A_1 \ltdyn A_2$ and
$c' : A_2 \ltdyn A_3$ and $x : A_1$ is related to $y : A_3$ via the composition
$c \comp c'$, then the upcast $\upc{c} x$ is related to $y$ via $c'$. This is
equivalent to the existence of the following square:

% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_1} \& {A_2} \& {A_3} \\
	{A_2} \&\& {A_3}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\upc{c}}"', from=1-1, to=2-1]
	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
	\arrow["\id", from=1-3, to=2-3]
	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
\end{tikzcd}\]

The three squares corresponding to the other rules are analogous. Observe that
although this is really a fact about the relationship between $c$ and $\upc{c}$,
we must universally quantify over all relations $c'$. This means we cannot
require the existence of this square as part of the definition of a monotone
relation, as it is self-referential. This would seem to imply that we cannot
give a syntax-independent semantic model for gradual typing. However, the key
observation is that through compositionality, we can \emph{derive} the above
squares from simpler ones that do not involve composition of relations. For
example, below is the simpler square corresponding to UpL:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_1} \& {A_2} \\
	{A_2} \& {A_2}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\upc{c}}"', from=1-1, to=2-1]
	\arrow["\id", from=1-2, to=2-2]
	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]

We can use this to derive the original UpL square by composing this square with
the identity square for $c'$:

% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwicihBXzIpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMSw1LCJcXGlkIiwyXV0=
\[\begin{tikzcd}[ampersand replacement=\&]
	{A_1} \& {A_2} \& {A_3} \\
	{A_2} \& {A_2} \& {A_3}
	\arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\upc{c}}"', from=1-1, to=2-1]
	\arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
	\arrow["\id"', from=1-2, to=2-2]
	\arrow["\id", from=1-3, to=2-3]
	\arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
	\arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
\end{tikzcd}\]

The existence of these simpler squares for a relation can be stated without
reference to other relations. Thus, we can require that these squares exist in
the definition of relation without running into the circularity issue we faced
with the original rules.

Finally, we use squares to model the term precision ordering $\Gamma \ltdyn
\Gamma' \vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the
monotone functions denoted by $M$ and $N$:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVUYgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVUZBJyJdLFsyLDMsIlxcc2Vte059Il0sWzAsMSwiXFxzZW17TX0iLDJdLFswLDIsIlxcR2FtbWFee1xcbHRkeW59IiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJVRmMiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dXQ==
\[\begin{tikzcd}[ampersand replacement=\&]
	\Gamma \& {\Gamma'} \\
	{UF A} \& {UFA'}
	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\sem{M}}"', from=1-1, to=2-1]
	\arrow["{\sem{N}}", from=1-2, to=2-2]
	\arrow["UFc"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]

@@

We have seen how we can recover transitive reasoning by working with the
lock-step error ordering. However, it turns out that we cannot completely
isolate the reasoning about steps from the reasoning about the ordering.
% Unfortunately, by defining the partial order on the guarded lift monad to be
% the lock-step error ordering, we introduce a new issue.
In particular, the cast rules required for proving graduality no longer hold if
interpreted using the lock-step ordering. The source of the problem is that the
downcast from the dynamic type to a function takes a step, i.e., involves a
$\theta$. Consider the DnL rule shown below:
%
\begin{mathpar}
  \inferrule*{c : A \ltdyn A' \and M \ltdyn N : A'}
             {\dnc{c}{M} \ltdyn N : c}
\end{mathpar}
%
Suppose that $c$ is $\iarr \colon \dyntodyn \ltdyn\, \dyn$, and recall the
definition of the semantics of a downcast given in Section
\ref{sec:term-interpretation}. The DnL rule corresponds to the following square:

>>>>

At this point we might try to weaken from our initial attempt at
providing a poset semantics to a semantics where types are equipped
with merely a reflexive relation, and continue using the
step-insensitive ordering. However some level of transitivity is
absolutely essential to providing compositional reasoning, and is used
pervasively in the New-Licata development.
%
The reason that it is essential has to do with the details of the
graduality proof, so first we expand on how we prove graduality
compositionally. First of all, we need to extend our compositional
semantics of types and terms to also a compositional semantics of
\emph{type precision derivations} and \emph{term precision}.
%
Sticking to a poset-based semantics, if our value types are posets, we then want
our precision derivations $c : A \ltdyn A'$ to be \emph{poset relations} on $A,
A'$, that is relations between the underlying sets of $A, A'$ that are
\emph{downward-closed} in $A$ and \emph{upward-closed} in $A'$.  We denote such
a relation between $A$ and $A'$ by $c : A \rel A'$. The obvious example of a
poset relation is the poset ordering itself, $\leq_A : A \rel A$.

% \max{TODO: Eric pick
%   up from here. Go into the squares for casts and how representability
%   gives us a compositional approach to proving them}

\eric{Picked up here; Max, please read this.}

To model term precision, we use the notion of \emph{square}, which specifies a
relation between two monotone functions. This has been used to model term
precision in previous work on double-categorical models for gradual typing
\cite{new-licata18}. The definition of square is as follows. Let $A_i, A_o,
A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o : A_o
\rel A_o'$ be poset relations, and let $f : A_i \to A_o$ and $g : A_i' \to A_o'$
be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x : A_i$
and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
visualize this situation as follows:
%
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBX2kiXSxbMCwxLCJBX28iXSxbMSwwLCJBX2knIl0sWzEsMSwiQV9vJyJdLFsyLDMsImciXSxbMCwxLCJmIiwyXSxbMCwyLCJjX2kiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFsxLDMsImNfbyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
  {A_i} \& {A_i'} \\
  {A_o} \& {A_o'}
  \arrow["{c_i}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["f"', from=1-1, to=2-1]
  \arrow["g", from=1-2, to=2-2]
  \arrow["{c_o}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
%

\eric{Not sure if this belongs here...}
Squares are used to model the term precision ordering $\Gamma \ltdyn \Gamma'
\vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the monotone
functions denoted by $M$ and $N$:

% https://q.uiver.app/#q=WzAsNCxbMCwwLCJcXEdhbW1hIl0sWzAsMSwiVVxcbGkgQSJdLFsxLDAsIlxcR2FtbWEnIl0sWzEsMSwiVVxcbGkgQSciXSxbMiwzLCJcXHNlbXtOfSJdLFswLDEsIlxcc2Vte019IiwyXSxbMCwyLCJcXEdhbW1hXntcXGx0ZHlufSIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzEsMywiVVxcbGkgYyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
	\Gamma \& {\Gamma'} \\
	{U\li A} \& {U\li A'}
	\arrow["{\Gamma^{\ltdyn}}", "\shortmid"{marking}, no head, from=1-1, to=1-2]
	\arrow["{\sem{M}}"', from=1-1, to=2-1]
	\arrow["{\sem{N}}", from=1-2, to=2-2]
	\arrow["{U\li c}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]

Having defined squares, we now discuss the interpretation of the cast rules UpL
and UpR \footnote{The cast rules DnL and DnR involve error domain relations and
squares, which for the sake of this discussion are not relevant.}. These rules
specify a relationship between the semantics of type precision and the
corresponding casts (see Section \ref{sec:GTLC}). For example, consider the UpL
rule. It states that if $c : A_1 \ltdyn A_2$ and $c' : A_2 \ltdyn A_3$ and $x :
A_1$ is related to $y : A_3$ via the composition $c \comp c'$, then the upcast
$\upc{c} x$ is related to $y$ via $c'$. This is equivalent to the existence of
the following square:
%
% https://q.uiver.app/#q=WzAsNSxbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMyw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
\[\begin{tikzcd}[ampersand replacement=\&]
  {A_1} \& {A_2} \& {A_3} \\
  {A_2} \&\& {A_3}
  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["{\upc{c}}"', from=1-1, to=2-1]
  \arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
  \arrow["\id", from=1-3, to=2-3]
  \arrow["{c'}"', "\shortmid"{marking}, no head, from=2-1, to=2-3]
\end{tikzcd}\]
%
Observe that although this is really a fact about the relationship between $c$
and $\upc{c}$, we must universally quantify over all relations $c'$. This means
we cannot require the existence of this square as part of the definition of a
monotone relation, as it is self-referential. This would seem to imply that we
cannot give a syntax-independent semantic model for gradual typing. However, the
key observation is that through compositionality, we can \emph{derive} the above
squares from simpler ones that do not involve composition of relations. These
simpler squares have been used in prior work for giving a compositional
semantics to gradual typing \cite{new-licata18}. Below is the square
corresponding to UpL:
%
% https://q.uiver.app/#q=WzAsNCxbMCwwLCJBXzEiXSxbMCwxLCJBXzIiXSxbMSwwLCJBXzIiXSxbMSwxLCJBXzIiXSxbMCwyLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwzLCJyKEFfMikiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJiYXJyZWQifSwiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDEsIlxcdXBje2N9IiwyXSxbMiwzLCJcXGlkIl1d
\[\begin{tikzcd}[ampersand replacement=\&]
  {A_1} \& {A_2} \\
  {A_2} \& {A_2}
  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["{\upc{c}}"', from=1-1, to=2-1]
  \arrow["\id", from=1-2, to=2-2]
  \arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
\end{tikzcd}\]
%
This square, and the corresponding one related to UpR, together express the
\emph{representability} of $c$ by the upcast morphism $\upc{c}$, which
intuitively says that $c$ is determined by the behavior of $\upc{c}$. We can use
this simpler square to derive the original UpL square by horizontally composing
with the identity square for $c'$:
%
% https://q.uiver.app/#q=WzAsNixbMCwwLCJBXzEiXSxbMSwwLCJBXzIiXSxbMiwwLCJBXzMiXSxbMCwxLCJBXzIiXSxbMiwxLCJBXzMiXSxbMSwxLCJBXzIiXSxbMCwxLCJjIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbMSwyLCJjJyIsMCx7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzMsNSwicihBXzIpIiwyLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoiYmFycmVkIn0sImhlYWQiOnsibmFtZSI6Im5vbmUifX19XSxbNSw0LCJjJyIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6ImJhcnJlZCJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV0sWzAsMywiXFx1cGN7Y30iLDJdLFsyLDQsIlxcaWQiXSxbMSw1LCJcXGlkIiwyXV0=
\[\begin{tikzcd}[ampersand replacement=\&]
  {A_1} \& {A_2} \& {A_3} \\
  {A_2} \& {A_2} \& {A_3}
  \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  \arrow["{\upc{c}}"', from=1-1, to=2-1]
  \arrow["{c'}", "\shortmid"{marking}, no head, from=1-2, to=1-3]
  \arrow["\id"', from=1-2, to=2-2]
  \arrow["\id", from=1-3, to=2-3]
  \arrow["{r(A_2)}"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  \arrow["{c'}"', "\shortmid"{marking}, no head, from=2-2, to=2-3]
\end{tikzcd}\]
%
The existence of these simpler squares for a relation can be stated without
reference to other relations, and hence we can require that these squares exist
in the definition of relation without running into the circularity issue we
faced with the original rules. This allows us to formulate a notion of model
that is independent of the particular syntax of any cast calculus.

We now see why compositionality of squares, and hence transitivity of the
relations on the objects that model types, is an essential property that we
cannot sacrifice.

@@

Given that compositional reasoning is essential, we seek to formulate an
ordering relation on $\li A$ that is transitive. The theorem above
tells us, however, that any such relation must either not be a congruence with
respect to $\theta$, or must not be left- and right-step-insensitive.
\footnote{Technically, we only need to give up step insensitivity on \emph{one}
side, as the above theorem requires $R$ to be both left- and
right-step-insensitive. We could define two separate ``step-semi-sensitive''
relations on $\li A$, each closed under delays on the left and
right respectively. The denotation of term precision would then involve the
conjunction of those relations. We do not pursue this approach in this paper,
because (1) it is not clear if the step-semi-sensitive relations would be
transitive, and (2) we would need to reason about the two ordering relations
separately for every pair of terms.}
%
We cannot forego the property of being a $\theta$-congruence, as without this we
would not be able to prove properties of the relation using \lob-induction,
e.g., that the monadic extension of a monotone function $f$ is a \emph{monotone}
function with respect to the ordering.
%and moreover this would mean that $\theta$ would not be a monotone function.

Thus, we choose to sacrifice left and right step-insensitivity. That is, we will
define an ordering relation that requires terms to be in ``lock-step''. In order
for two computations to be related in this ordering, they must have the same
stepping behavior (or the left-hand side must be an error). To deal with terms
that take differing numbers of steps, we then define a separate relation called
\emph{weak bisimilarity} that relates terms that are extensionally equal and may
differ by a finite number of delays. Then, the semantics of the error ordering
for the guarded lift monad will involve a combination of these two relations:
a ``closure'' of the lock-step error ordering under weak bisimilarity on both
sides.
% Although the combined relation will not be transitive (for the same reason that
% $\semltbad$ is not transitive), this...
This decomposition has the advantage that we can recover some transitive
reasoning and push much of the reasoning about stepping to the margins of the
development.

More formally, we define the \emph{lock-step error-ordering}, with the idea
being that two computations $l$ and $l'$ are related if they are in lock-step
with regard to their intensional behavior, up to $l$ erroring. Figure
\ref{fig:lock-step-error-ordering} gives the definition.

\begin{figure}
    \fbox{$l_1 \ltls l_2$}
    \begin{align*}
        &\eta\, x \ltls \eta\, y \text{ if } 
            x \mathbin{\le_A} y \\
        %		
        &\mho \ltls l' \\
        %
        &\theta\, \tilde{l} \ltls \theta\, \tilde{l'} \text{ if } 
            \later_t (\tilde{l}_t \ltls \tilde{l'}_t)
    \end{align*}
    \caption{The lock-step error ordering.}
    \label{fig:lock-step-error-ordering}
\end{figure}

When both sides are $\eta$, then we check that the returned values are related
in $\le_A$. The error term $\mho$ is the least element. Lastly, if both sides step
(i.e., are a $\theta$) then we compare the resulting computations one time step
later.
%
It is straightforward to prove using \lob-induction that this relation is
reflexive, transitive and anti-symmetric given that the underyling relation $R$
has those properties. The lock-step ordering is therefore the partial ordering
we will associate with $\li A$.
%
We similarly define a heterogeneous version of this ordering between the lifts
of two different types $A$ and $A'$, parameterized by a poset relation $c : A \rel A'$.

The second relation we define on the guarded lift monad is \emph{weak bisimilarity}.
% For a type $X$, we define a relation on $\li X$, called ``weak bisimilarity",
% written $l \bisim l'$. 
Intuitively, we say $l \bisim l'$ if they are the same ``up to delays''.
%
Weak bisimilarity on $\li A$ is parameterized by an underlying relation $R$ on $A$
and is defined by guarded fixpoint as shown in Figure \ref{fig:weak-bisimilarity}.

@@

@@

@@

\eric{We may want to introduce error domain relations and squares at this point,
since the squares for DnL and DnR involve lift.}

We have seen how we can recover transitive reasoning by working with the
lock-step error ordering. However, it turns out that we cannot completely
isolate the reasoning about steps from the reasoning about the ordering.
% Unfortunately, by defining the partial order on the guarded lift monad to be
% the lock-step error ordering, we introduce a new issue.
In particular, the cast rules required for proving graduality no longer hold if
interpreted using the lock-step ordering. The source of the problem is that the
downcast from the dynamic type to a function takes a step, i.e., involves a
$\theta$, and the lock-step ordering requires both sides to be in sync. More
specifically, consider the DnL rule for $\iarr \colon \dyntodyn \ltdyn\, \dyn$
shown below:
%
\begin{mathpar}
  \inferrule*{M \ltdyn N : \iarr}
             {\dnc{\iarr}{M} \ltdyn N : c}
\end{mathpar}
%
This rule corresponds to the following square:
